---
title: "Multiple Views"
subtitle: "INFO 526 Data Analysis and Visualization"
author: "Adriana Picoral"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The temperature of emotions

We will be open science experimental data on [the temperature of emotions](https://osf.io/tcbg5/) for this part of the tutorial. I downloaded the csv file from the osf.io website and placed it inside the data folder in my project. As usual, we will load tidyverse and read the data in.

```{r}
library(tidyverse)
emotion_temperature <- read_csv("data_raw/module 3 data/dte_exp1.csv")
glimpse(emotion_temperature)
```
## Data Inspection

Let's do a quick inspection on the categorical variables in the data. In addition to `count()` I will be using `kable()` to make the tables look nice in the output document.

### How many observations per language do we have?

```{r}
# create table with counts of data points per participant language, sort table
# since language is an unordered categorical variable
emotion_temperature %>%
  count(UserLanguage, sort = TRUE) %>%
  kable(caption = "Number of observations per language group")
```

### How many participants per language group?

We have multiple observations per participant, so we need to first ensure we extract distinct participant IDs per language group to then count observations.

```{r}
emotion_temperature %>%
  distinct(rid, UserLanguage) %>%
  count(UserLanguage, sort = TRUE) %>%
  kable(caption = "Number of participants per language group")
```

### What are the demographics of the participants, besides language?

To get more in depth on participants' demographics, let's create a dataframe with distinct info related to each participant.

```{r}
participant_info <- emotion_temperature %>%
  distinct(rid, UserLanguage, firstlang, age, gender, country)

participant_info %>%
  head() %>%
  kable(caption = "First six participants, and their demographic information")
```

Participant IDs should not repeat in the table above. Let's check that that's actually the case.

```{r}
# are there any repeated participant IDs? Count IDs and filter to see if any
# ID has count higher than 1
participant_info %>%
  count(rid, sort = TRUE) %>%
  filter(n > 1)
```

No repeated IDs, so everything looks good, we might proceed with our `participant_info` data analysis.

### What is the average participant age?

```{r}
participant_info %>% 
  group_by(country) %>%
  summarize(mean_age = mean(age)) %>%
  kable(caption = "Mean participant age per gender group",
        digits = 0)

review(emotion_temperature)
```

Now that we have a general idea of who the participants are (and I encourage you to explore the other categorical variables in the data in a similar way), we will proceed to the actual experimental data.

## Visualizing responses by language and emotion

To better visualize the value of `response` across different categorical variables, we will be using multiple views to split our plot into subplots. 

If you are curious about the work that goes into creating palettes, you can watch [this conference presentation on the viridis palette](https://youtu.be/xAoljeRJ3lU?t=1032).

```{r}
library(tidytext)
emotion_temperature %>%
  group_by(emotion, UserLanguage)  %>% 
  summarize(mean_response = mean(response)) %>% 
  ggplot(aes(y = reorder_within(emotion, mean_response, UserLanguage),
             x = mean_response,
             color = mean_response)) +
  geom_point() +
  facet_wrap(~UserLanguage, scales = "free_y") +
  #scale_color_continuous(type = "viridis") +
  #scale_color_gradient(low = "blue", high = "red") +
  scale_color_gradientn(colours = hcl.colors(5, "Berlin")) +
  scale_y_reordered() +
  theme(legend.position = "none") +
  labs(y = "",
       title = "Average response per emotion across language group")

```

```{r fig.height=10}
emotion_temperature %>%
  filter(gender != "Other") %>%
  group_by(emotion, UserLanguage, gender)  %>% 
  summarize(mean_response = mean(response)) %>% 
  ggplot(aes(y = UserLanguage,
             x = mean_response,
             color = mean_response)) +
  geom_point() +
  facet_grid(emotion~gender, scales = "free_y") +
  #scale_color_continuous(type = "viridis") +
  #scale_color_gradient(low = "blue", high = "red") +
  scale_color_gradientn(colours = hcl.colors(5, "Berlin")) +
  theme(legend.position = "none",
        strip.text = element_text(size = 5)) +
  labs(y = "",
       title = "Average response per emotion across language group")
```

