---
title: "Scales"
subtitle: "INFO 526 Data Analysis and Visualization"
author: "Adriana Picoral"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# TV show ratings

We will be working with two [subsets of IMDb data](https://www.imdb.com/interfaces/) that are available for access to customers for personal and non-commercial use.

`title.episode.tsv` contains the tv episode information. Fields include:

* tconst (string) - alphanumeric identifier of episode
* parentTconst (string) - alphanumeric identifier of the parent TV Series
* seasonNumber (integer) – season number the episode belongs to
* episodeNumber (integer) – episode number of the tconst in the TV series

`title.ratings.tsv` contains the IMDb rating and votes information for titles. Fields include:

* tconst (string) - alphanumeric unique identifier of the title
* averageRating – weighted average of all the individual user ratings
* numVotes - number of votes the title has received

We first read both files in, which I downloaded from the [IMDb website](https://datasets.imdbws.com/) and added to my `data` folder in my project.

```{r}
library(tidyverse)
imdb_ratings <- read_tsv("data/title.ratings.tsv")
series_data <- read_tsv("data/title.episode.tsv")
```

Then we merge the files with `left_join()` using the `series_data` data frame as the reference.

```{r}
series_data_complete <- series_data %>%
  left_join(imdb_ratings)
```

The data is very messy, we will be looking at the variables `episodeNumber` and `seasonNumber` so we will parse these variables as number and limit our observations to a seasons up to 30 and episode up to 20. This will make the data more manageable for visualization.

```{r}
series_data_clean <- series_data_complete %>%
  mutate(episodeNumber = parse_number(episodeNumber),
         seasonNumber = parse_integer(seasonNumber)) %>%
  filter(episodeNumber > 0 & episodeNumber < 21 &
           seasonNumber < 31 & !is.na(averageRating &
                                        numVotes > 0))
```

## Multiple views and scale transformation

We will be using the `patchwork` library here to compare two plots of the same data side-by-side. Instead of just outputting the plots directly, we will save them as objects first and then combine them.

```{r}
library(patchwork)

# create first plot, save it to an object
original_plot <- series_data_clean %>%
  ggplot(aes(x = averageRating,
             y = numVotes)) +
  geom_point(alpha = .3) +
  labs(title = "Original number of votes")

# created plot with transformed axis, save it to an object
transformed_plot <- series_data_clean %>%
  ggplot(aes(x = averageRating,
             y = numVotes)) +
  geom_point(alpha = .3) +
  scale_y_log10() +
  labs(title = "Log-transformed number of votes")

# output both plots side by side (need patchwork library for this to work)
original_plot + transformed_plot
```

The idea behind log transforming a variable is to display a nonlinear relationship as a linear relationship, so plots are easier to read. Log transformation is often apply to count data (like our `numVotes` variable in this data set), which is often skewed. 

## Mean rating across seasons and episodes

The hypothesis here is that quality of episodes overall decreases over time. In other words, for later seasons and later episodes, rating will be lower.

Let's create aggregated data by season and episode, by calculating the mean of average rating. We will make sure to remove NAs with `na.rm = TRUE` (alth.

```{r}
episode_season_rating <- series_data_clean  %>%
  group_by(seasonNumber, episodeNumber) %>%
  summarize(mean_rating = mean(averageRating, na.rm = TRUE))
```

We can now visualize our aggregated data on a heat map with the viridis color scheme.

```{r}
episode_season_rating %>%
  ggplot(aes(x = seasonNumber,
             y = episodeNumber,
             fill = mean_rating)) +
  geom_tile() +
  scale_fill_continuous(type = "viridis") 
```

We can also visualize the same data with a scatterplot instead of a heatmap.

```{r}
episode_season_rating %>%
  ggplot(aes(x = seasonNumber,
             y = mean_rating,
             color = episodeNumber)) +
  geom_point() +
  geom_point(shape = 21,
             color = "black",
             alpha = .3) +
  scale_color_gradientn(colours = hcl.colors(5, palette = "Grays", rev = TRUE)) +
  theme_linedraw() 
```

Note the y scale in the plot above -- it has a very limited range, and one could argue that displaying the full range of possible ratings (from 1 to 10) is a more accurate representation of the rating variation. In the case of the plot above, variation is represented less dramatically, since the the larger y scale flattens the differences in ratings across season number.

```{r}
episode_season_rating %>%
  ggplot(aes(x = seasonNumber,
             y = mean_rating,
             color = episodeNumber)) +
  geom_point() +
  geom_point(shape = 21,
             color = "black",
             alpha = .3) +
  scale_color_gradientn(colours = hcl.colors(5, palette = "Grays", rev = TRUE)) +
  theme_linedraw() +
  ylim(1,10)
```
```


