---
title: "Annotating Plots"
subtitle: "INFO 526 Data Analysis and Visualization"
author: "Adriana Picoral"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

We will focus on how to use text on this tutorial to annotate a timeseries. We will be getting data from two sources: annotation data will be scraped from [a news website](https://abcnews.go.com/Business/gamestop-timeline-closer-saga-upended-wall-street/story?id=75617315) using the `revest` package, and stock closing prices for GameStop will be downloaded from [Yahoo! Finance](https://finance.yahoo.com/quote/GME/history?p=GME).

Our visualization will be of GameStop closing prices for their stock between December 2020 and March 2021.

# Scraping text data from a website

We will start with scraping data from [a website which contains information on the timeline for the changes in GameStop stock prices](https://abcnews.go.com/Business/gamestop-timeline-closer-saga-upended-wall-street/story?id=75617315).

The first step is to read the html page in.

```{r}
# load libraries
library(tidyverse)
library(rvest)

# scrape text from news website
# set URL
article_url <- "https://abcnews.go.com/Business/gamestop-timeline-closer-saga-upended-wall-street/story?id=75617315"

# read the URL into an object that holds the html code for the page
article_html <- read_html(article_url)
```

Once we have the HTML page, we can parse for HTML nodes. For this specific page, the information we want is in `<h3>` tags.

```{r}
article_html %>%
  html_nodes("h3") 
```

We can add to this by parsing the text in these HTML tags.

```{r}
article_html %>%
  html_nodes("h3") %>%
  html_text()
```

If we transform the output into a dataframe, we can then transform it.

```{r}
# save character vector to an object
raw_text <- article_html %>%
  html_nodes("h3") %>%
  html_text()

# create a dataframe with raw text
gme_annotation <- data.frame(original_text = raw_text)
```

We can split the date from the event text.

```{r}
# split date and annotation
gme_annotation <- gme_annotation %>%
  separate(original_text, sep = ":",
           into = c("date", "event"))

gme_annotation
```

We can filter out the line that has `NA` for event.

```{r}
gme_annotation <- gme_annotation %>%
  filter(!is.na(event)) 
```

Then we focus on the date. We first inspect it and try to convert it to a date data format.

```{r}
gme_annotation %>%
  pull(date)

gme_annotation %>%
  mutate(date = gsub("\\.", "", date),
         updated_date = parse_datetime(date,
                               format = "%b %d, %Y")) %>%
  select(date, updated_date)
```

Looks good, so we can make the change to the actual dataframe.

```{r}
gme_annotation <- gme_annotation %>%
  mutate(date = gsub("\\.", "", date),
         date = parse_datetime(date,
                               format = "%b %d, %Y"))
```

We have everything we need for our annotation.

# Getting data from stock prices

For the stock prices, we will get data from [Yahoo](https://finance.yahoo.com/quote/GME/history?p=GME). We will use `janitor` to `clean_names()` and lowercase the column names from the original data.

```{r}
library(janitor)
gme_data <- read_csv("data/GME.csv") %>%
  clean_names()
```

We can now merge the stock prices dataframe with annotation dataframe, since both have a column called `date` that is a date format.

```{r}
# merge dataframes, using the largest dataframe as the reference (i.e. left)
gme_data <- gme_data %>%
  left_join(gme_annotation)
```

# Visualization

We will plot our stock price data on a line using `geom_line()`, with points only for the data we have annotations for. We will annotate the timeseries using `geom_label()` but from the `ggrepel` package. We will also use `annotate()`. Because of the number of annotations labels (11 total), we need to create a custom color scheme. The x scale is adjusted with `filter` and the y axis is set for a dollar format.

```{r}
library(ggthemes)
library(ggrepel)
library(RColorBrewer)
library(scales)

my_color_scheme <- c(brewer.pal(n = 8, name = "Dark2"),
                     rev(brewer.pal(n = 3, name = "Dark2")))

gme_data %>%
  filter(date > as.Date("2020-12-01") &
           date < as.Date("2021-03-01")) %>%
  ggplot(aes(x = date,
             y = close,
             label = event,
             color = event)) +
  geom_line(color = "black") +
  geom_point(data = gme_data %>% filter(!is.na(event))) +
  geom_label_repel(size = 1.8,
                   nudge_y = 200) + 
  scale_y_continuous(limits = c(0, 600),
                     labels = scales::dollar_format(),
                     breaks = c(0, 100, 200, 300, 350)) +
  #scale_color_brewer(type = "qual", palette = "Paired") +
  scale_color_manual(values = my_color_scheme) +
  geom_hline(yintercept = 347.51,
             color = "dimgrey",
             linetype = "dashed") +
  annotate("text",
           x = as.POSIXct("2020-12-20"),
           y = 375,
           color = "dimgrey",
           label = "Stock Closing High: $347.51") +
  theme_linedraw() +
  theme(panel.grid = element_blank(),
        legend.position = "none") +
  labs(title = "GameStop Stock Closing Price Fluctuation",
       subtitle = "between December 2020 and March 2021",
       caption = "data from finance.yahoo.com and abcnews.go.com") 
```

