---
title: "Gestalt Principles"
subtitle: "INFO 526 Data Analysis and Visualization"
author: "Adriana Picoral"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Gestalt Principles

We humans are extremely good at finding patterns, even when no pattern exists. This ability to perceive different shapes and lines as a unified image rather than just a bunch of elements that are not related to each other follow what we call *Gestalt* principles (Healy, 2018). These principles state that visual elements are perceived as being related to each other when these elements are:

1. close to each other -- i.e., *Proximity* principle
1. similar in shape, color and size -- i.e., *Similarity* principle
1. visually tied -- i.e., *Connection* principle
1. perceived as being part paths, lines, curves, even when some elements are "hidden" -- i.e., *Continuity* principle
1. not complete; we construct the incomplete form into familiar shapes  -- i.e., *Closure* principle
1. perceived to be either in the foreground or in the background together  -- i.e., *Figure and Ground* principle
1. perceived are sharing a direction of movement -- i.e., *Common fate* principle


Let's first create simulation data, with 4 different groups, to create a plot that illustrates the principles above.

```{r}
library(tidyverse)
# create simulation data
my_circles <- data.frame(y = rep(c(1:4), 8), 
                         x = c(rep(1, 4),
                               rep(2, 4),
                               rep(3, 4),
                               rep(4, 4),
                               rep(5, 4),
                               rep(6, 4),
                               rep(7, 4),
                               rep(8, 4))) %>%
  mutate(group = ifelse(y %% 2 == 0,
                        "A", "B"),
         group_shape = ifelse(x %% 2 == 0,
                        "A", "B"),
         group_line = case_when(y == 1 ~ "A",
                                y == 2 ~ "B",
                                y == 3 ~ "C",
                                y == 4 ~ "D"),
         x = x/2,
         y = y/2)
```

Figure 1 shows how elements are perceived to be grouped different according to different principles.

```{r gestalt, fig.cap="Grouping is done by different principles: proximity, similarity (i.e., color and shape, and connection).", fig.height=12}
# first plot with circles with no color
p1 <- my_circles %>%
  filter(x != 3) %>%
  ggplot(aes(x = x,
             y = y)) +
  geom_point(size = 10, color = "white") +
  geom_point(size = 10,
         color = "black",
         fill = "white",
         stroke = 2,
         shape = 1) +
  theme_minimal() +
  ylim(-1, 2.1) +
  xlim(-1, 4.1) +
  labs(x = "", y = "") +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c("white", "royalblue3"))

# second plot uses color to establish groupings
p2 <- my_circles %>%
  filter(x != 3) %>%
  ggplot(aes(x = x,
             y = y)) +
  geom_point(size = 10,
         aes(color = group)) +
  geom_point(size = 10,
         color = "black",
         fill = "white",
         stroke = 2,
         shape = 1) +
  theme_minimal() +
  ylim(-1, 2.1) +
  xlim(-1, 4.1) +
  labs(x = "", y = "") +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c("white", "royalblue3"))

# third plot links circles for grouping data points horizontally
p3 <- my_circles %>%
  filter(x != 3) %>%
  ggplot(aes(x = x,
             y = y)) +
  geom_line(aes(group = group_line)) +
  geom_point(size = 10,
            color = "white") +
  geom_point(size = 10,
         color = "black",
         fill = "white",
         shape = 1) +
  theme_minimal() +
  ylim(-.1, 2.1) +
  xlim(-1, 4.1) +
  labs(x = "", y = "") +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c("white", "royalblue3"))

# fourth plot uses shaping for grouping
p4 <- my_circles %>%
  filter(x != 3) %>%
  ggplot(aes(x = x,
             y = y)) +
  geom_point(size = 10,
         color = "black",
         fill = "white",
         aes(shape = group_shape)) +
  theme_minimal() +
  ylim(-1, 2.2) +
  xlim(-1, 4.1) +
  labs(x = "", y = "") +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c("white", "royalblue3"))


# display all plots together (need patchwork library for this)
library(patchwork)
p1 / p2 / p3 / p4
```